---
title: "spaceMarker example"
date: "`r format(Sys.time(), '%d %b %Y')`"
vignette: >
    %\VignetteIndexEntry{spaceMarker example}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
output: rmarkdown::html_vignette
html_document:
    fig_caption: yes
    fig_retina: FALSE
    keep_md: FALSE
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Introduction

The dataset used in this analysis comprises a select subset from two 
replicates of Xenium human breast cancer tissue Sample 1. This subset was 
extracted from the raw dataset as described in the R script
located at /inst/generate_vignette_data.R.

```{r}
library(Seurat)
library(ggplot2)
library(spaceMarker)
library(data.table)
library(dplyr)
library(glmnet)
library(caret)
library(corrplot)
library(igraph)
library(ggraph)
library(ggrepel)
library(gridExtra)
library(utils)
library(spatstat)
```

# Load data 
```{r load subsection of rep1}
data(rep1_sub, rep1_clusters, rep1_neg)
rep1_clusters$cluster=factor(rep1_clusters$cluster,levels=paste("c",1:8, sep=""))
```
The requried input data structure for the main function `lasso_markers()` 
is a list of matrices. For loading the 10x Xenium data, the function
`get_data()` can be used. If you are working with Merscope or CosMx data, 
the following code can be used for constructing the required data structure. 

### merscope  
The cosntructed merscope_data can be used in function `get_vectors()`.
The defined nc_lst can be used in function` create_genesets()`. 
```{r merscope example, eval=FALSE}
# count matrix
# rows as genes and columns as cells
# gene name as row names 
data_p = "path/to/merscope/output/data/"
cm = qread("./counts.qs")
# cell label and coordinates 
clusters_info =  read.csv("./cells_meta.csv")

# -----------------------------------------------------------------------------
# if transcript coordinates are available, need to repeat for each sample
# load transcript data, transcripts must contain follwing columns 
# - "x": x coordinate
# - "y": y coordinate 
# - "feature_name": gene name 
tr_req_cols = c("x", "y", "feature_name")
transcript_df =  read.csv("./transcripts.csv")
merscope_data = list(cm = cm, trans_info = transcript_df[,req_cols])
# the defined merscope_data can be used in function get_vectors() 

# the cluster label, sample label and coordinates for each cell 
cl_req_cols = c("x","y","cluster","sample")
clusters_df = clusters_info[,cl_req_cols]

# -----------------------------------------------------------------------------
# define negative control objects if available, need to repeat for each sample
# load coordinates for each detection, must contain following columns 
# - "x": x coordinate
# - "y": y coordinate 
# - "target": negative control names 
# - "sample": sample label 
nc_coords = read.csv("./negative_controls_detections.csv")
nc_names = unique(nc_coords$feature_name)
kpt_cols = c("x","y","feature_name","sample")
nc_lst = list(trans_info = nc_coords[,kpt_cols])



```

### CosMx
The constructed cosmx_data can be used in function `get_vectors()`.
The defined nc_lst can be used in function` create_genesets()`. 
```{r cosmx example, eval=FALSE}
tiledbURI <- "file/path/to/TileDB/array/on/local/machine"

# read in SOMACollection
tiledb_scdataset <- tiledbsc::SOMACollection$new(uri = tiledbURI, verbose = FALSE)
cm <- tiledb_scdataset$somas$RNA$X$members$counts$to_matrix(batch_mode = TRUE)

# the cluster label, sample label and coordinates for each cell 
clusters_info <- tiledb_scdataset$somas$RNA$obs$to_dataframe()

# -----------------------------------------------------------------------------
# if transcript coordinates are available, need to repeat for each sample
# load transcript data, transcripts must contain follwing columns 
# - "x": x coordinate
# - "y": y coordinate 
# - "feature_name": gene name 
req_cols = c("x","y","feature_name")
cosmx_data = list(cm = cm, trans_info = transcript_df[,c(req_cols)])
# 

# clusters and cell locations 
cl_req_cols = c("x","y","cluster","sample")
clusters_df = clusters_info[,cl_req_cols]

# -----------------------------------------------------------------------------
# if available
# define negative control objects if available
# need to repeat for each background category and each sample
# load coordinates for each detection, must contain follwing columns 
# - "x": x coordinate
# - "y": y coordinate 
# - "target": negative control names 
# - "sample": sample label 
nc_coords = read.csv("./negative_controls_detections.csv")
nc_names = unique(nc_coords$feature_name)
kpt_cols = c("x","y","feature_name","sample")
nc_lst = list(trans_info = nc_coords[,kpt_cols])
# 



```

If you have a Seurat object, you can build the required object as follows 
instead of using the `get_data` function. The defined example_vectors_cm/
example_vectors_tr can be passed to `lasso_markers` to identify marker genes. 
## from a Seurat object 
```{r eval=FALSE}
seu_obj =Seurat::CreateSeuratObject(counts = rep1_sub$cm)
# make sure the clusters information contains column names: 
# cluster, x, y and sample
clusters_info = rep1_clusters
# make sure the transcript information contains column names: 
# feature_name, x, y
transcript_coords = rep1_sub$trans_info
data_example = list(cm=seu_obj@assays$RNA$counts,
                    trans_info =transcript_coords)

w_x =  c(min(floor(min(data$trans_info$x)), floor(min(clusters_info$x))), 
        max(ceiling(max(data$trans_info$x)), ceiling(max(clusters_info$x))))
w_y =  c(min(floor(min(data$trans_info$y)), floor(min(clusters_info$y))), 
        max(ceiling(max(data$trans_info$y)), ceiling(max(clusters_info$y))))

# build spatial vectors from transcript coordinates and cluster coordinates 
example_vectors_tr = get_vectors(data_lst= list(data_example),
                                cluster_info = clusters_info,
                                bin_type="square",
                                bin_param=c(10,10), 
                                all_genes = row.names(data_example$cm), 
                                w_x=w_x, w_y=w_y)
# -----------------------------------------------------------------------------
# build spatial vectors from count matrix and cluster coordinates 
# make sure the cluster information contains column names: 
# cluster, x, y, sample and cell_id
colnames(clusters_info)[5] = "cell_id"
example_vectors_cm = get_vectors(data_lst= NULL,
                                cm_lst=list(rep1=seu_obj@assays$RNA$counts),
                                cluster_info = clusters_info,
                                bin_type="square",
                                bin_param=c(10,10), 
                                all_genes = row.names(data_example$cm), 
                                w_x=w_x, w_y=w_y)

```

## Visualise the clusters
We can plot the coordinates of cells for each cluster of Replicate 1,
```{r rep clusters vis, fig.height=5,  fig.width=8,warning=FALSE}
p1<-ggplot(data = rep1_clusters,
        aes(x = x, y = y, color=cluster))+
        geom_point(position=position_jitterdodge(jitter.width=0,
                                        jitter.height=0), size=0.1)+
        scale_y_reverse()+
        theme_classic()+
        facet_wrap(~sample)+
        scale_color_manual(values = c("#FC8D62","#66C2A5" ,"#8DA0CB","#E78AC3",
                        "#A6D854","skyblue","purple3","#E5C498"))+
        guides(color=guide_legend(title="cluster", nrow = 2,
                        override.aes=list(alpha=1, size=2)))+
        
        theme(axis.text.x=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks=element_blank(),
                axis.title.x=element_blank(),
                axis.title.y=element_blank(),
                panel.spacing = unit(0.5, "lines"),
                legend.position="none",
                strip.text = element_text(size = rel(1)))+
        xlab("")+
        ylab("")
p2<-ggplot(data = rep1_clusters,
        aes(x = x, y = y, color=cluster))+
        geom_point(position=position_jitterdodge(jitter.width=0, 
                                    jitter.height=0), size=0.1)+
        facet_wrap(~cluster, nrow = 2)+
        scale_y_reverse()+
        theme_classic()+
        scale_color_manual(values = c("#FC8D62","#66C2A5" ,"#8DA0CB","#E78AC3",
                        "#A6D854","skyblue","purple3","#E5C498"))+
        guides(color=guide_legend(title="cluster", nrow = 1,
        override.aes=list(alpha=1, size=10)))+
        theme(axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.title.x=element_blank(),
            axis.title.y=element_blank(),
            panel.spacing = unit(0.5, "lines"), 
            legend.text = element_text(size=10),
            legend.position="top",
            legend.title = element_text(size=10),
            strip.text = element_text(size = rel(1)))+
        xlab("")+
        ylab("")

spacer <- patchwork::plot_spacer()
layout_design <- (p1 / spacer) | p2

layout_design <- layout_design + 
                 patchwork::plot_layout(widths = c(1, 4), heights = c(1, 1)) 

print(layout_design)
```

# Spatial vectors
We will create spatial vectors to convert coordinates to a
one-dimensional vector for every gene and every cluster. 
These spatial vectors may take the form of squares, rectangles, or hexagons. 

### Visualise the spatial vectors for genes
```{r plot spatial genes, , fig.width=3, fig.height=3}
w_x =  c(min(floor(min(rep1_sub$trans_info$x)),
        floor(min(rep1_clusters$x))), 
        max(ceiling(max(rep1_sub$trans_info$x)),
        ceiling(max(rep1_clusters$x))))
w_y =  c(min(floor(min(rep1_sub$trans_info$y)),
        floor(min(rep1_clusters$y))), 
        max(ceiling(max(rep1_sub$trans_info$y)),
        ceiling(max(rep1_clusters$y))))

loc_mt = as.data.frame(rep1_sub$trans_info[rep1_sub$trans_info$feature_name %in% 
          c("ERBB2","EPCAM"),
          c("x","y","feature_name")]%>%distinct())
colnames(loc_mt)=c("x","y","feature_name")

to_plot = c("ERBB2","EPCAM")

# plot square binning 
for (nm in to_plot){
    curr<-loc_mt[loc_mt[,"feature_name"]==nm,c("x","y")] %>%distinct()
    curr_ppp <- ppp(curr$x,curr$y,w_x, w_y)
    vec_quadrat <- quadratcount(curr_ppp, 10,10)
    title_nm <- paste(paste(nm, "square binning", sep=" - "))
    #plot(vec_quadrat, main= title_nm)
    vec_its <- intensity(vec_quadrat, image=TRUE)
    plot(vec_its, main = title_nm)  

}  

# plot hex binning 
for (nm in to_plot){
    w <- owin(xrange=w_x, yrange=w_y)
    H <- hextess(W=w, 50)
    bin_length <- length(H$tiles)
    curr<-loc_mt[loc_mt[,"feature_name"]==nm,c("x","y")] %>%distinct()
    curr_ppp <- ppp(curr$x,curr$y,w_x, w_y)
    
    title_nm <- paste(paste(nm, "hex binning", sep=" - "))
    vec_quadrat <- quadratcount(curr_ppp, tess=H)
    # plot(vec_quadrat, main= title_nm)
    vec_its <- intensity(vec_quadrat, image=TRUE)
    plot(vec_its, main = title_nm)  
   
}  

```

### Visualise the spatial vectors for clusters
```{r plot spatial clusters, fig.width=3, fig.height=3}

w_x =  c(min(floor(min(rep1_sub$trans_info$x)),
            floor(min(rep1_clusters$x))), 
        max(ceiling(max(rep1_sub$trans_info$x)),
            ceiling(max(rep1_clusters$x))))
w_y =  c(min(floor(min(rep1_sub$trans_info$y)),
            floor(min(rep1_clusters$y))), 
        max(ceiling(max(rep1_sub$trans_info$y)),
            ceiling(max(rep1_clusters$y))))

loc_mt =as.data.frame(rep1_clusters[rep1_clusters$cluster %in% c("c1","c2"),
            c("x","y","cluster")])

colnames(loc_mt)=c("x","y","cluster")
to_plot = c("c1","c2")
# plot square binning 
for (nm in to_plot){
    curr<-loc_mt[loc_mt[,"cluster"]==nm, c("x","y")]%>%distinct()
    curr_ppp <- ppp(curr$x,curr$y,w_x, w_y)
    vec_quadrat <- quadratcount(curr_ppp, 10,10)
    title_nm <- paste(paste(nm, "square binning", sep=" - "))
    #plot(vec_quadrat, main= title_nm)
    vec_its <- intensity(vec_quadrat, image=TRUE)
    plot(vec_its, main = title_nm)  

}  

# plot hex binning 
for (nm in to_plot){
    w <- owin(xrange=w_x, yrange=w_y)
    H <- hextess(W=w, 50)
    bin_length <- length(H$tiles)
    curr<-loc_mt[loc_mt[,"cluster"]==nm,c("x","y")] %>%distinct()
    curr_ppp <- ppp(curr$x,curr$y,w_x, w_y)
    
    title_nm <- paste(paste(nm, "hex binning", sep=" - "))
    vec_quadrat <- quadratcount(curr_ppp, tess=H)
    # plot(vec_quadrat, main= title_nm)
    vec_its <- intensity(vec_quadrat, image=TRUE)
    plot(vec_its, main = title_nm)  
   
}  

```

The function `get_vectors()` can be used to create spatial vectors for
all the genes and clusters.

### Spatial vectors for all genes and clusters
```{r create rep1 all vectors}
all_genes = row.names(rep1_sub$cm)
seed_number= 589

w_x =  c(min(floor(min(rep1_sub$trans_info$x)),
            floor(min(rep1_clusters$x))), 
        max(ceiling(max(rep1_sub$trans_info$x)),
            ceiling(max(rep1_clusters$x))))
w_y =  c(min(floor(min(rep1_sub$trans_info$y)),
            floor(min(rep1_clusters$y))), 
        max(ceiling(max(rep1_sub$trans_info$y)),
            ceiling(max(rep1_clusters$y))))


# get spatial vectors
rep1_sq10_vectors = get_vectors(data_lst= list(rep1_sub),
                                cluster_info = rep1_clusters,
                                bin_type="square",
                                bin_param=c(10,10), all_genes = all_genes , 
                                w_x=w_x, w_y=w_y)
```

The constructed spatial vectors can be used to quantify cluster-cluster and
gene-gene correlation.
### Cluster-Cluster correlation
```{r fig.width=6, fig.height=6}
exp_ord = paste("c", 1:8, sep="")
rep1_sq10_vectors$cluster_mt = rep1_sq10_vectors$cluster_mt[,exp_ord]
cor_cluster_mt = cor(rep1_sq10_vectors$cluster_mt,
                rep1_sq10_vectors$cluster_mt, method = "pearson")
# Calculate pairwise correlations
cor_gene_mt = cor(rep1_sq10_vectors$gene_mt, rep1_sq10_vectors$gene_mt,
                    method = "pearson")

col <- grDevices::colorRampPalette(c("#4477AA", "#77AADD", 
                                    "#FFFFFF","#EE9988", "#BB4444"))

corrplot::corrplot(cor_cluster_mt, method="color", col=col(200), diag=TRUE,
                addCoef.col = "black",type="upper",
                tl.col="black", tl.srt=45, mar=c(0,0,5,0),sig.level = 0.05, 
                insig = "blank", 
                title = "cluster-cluster correlation (square bin = 40x40)"
                )
```

### Gene-Cluster correlation
```{r gene_cluster_corr, warning=FALSE, message=FALSE,fig.width=6,fig.height=6}

cor_genecluster_mt = cor(x=rep1_sq10_vectors$gene_mt, 
                        y=rep1_sq10_vectors$cluster_mt, method = "pearson")

gg_correlation = as.data.frame(cbind(apply(cor_gene_mt, MARGIN=1, 
                                            FUN = mean, na.rm=TRUE),
                                        apply(cor_genecluster_mt, MARGIN=1, 
                                            FUN = mean, na.rm=TRUE)))
colnames(gg_correlation) = c("mean_correlation","mean_cluster")  
gg_correlation$gene=row.names(gg_correlation)

plot(ggplot(data = gg_correlation, 
    aes(x= mean_correlation, y=mean_cluster))+
        geom_point()+
        geom_text_repel(aes(label=gg_correlation$gene), size=1.8, hjust=1)+
    theme_bw()+
    theme(legend.title=element_blank(),
            axis.text.y = element_text(size=20),
            axis.text.x = element_text(size=20),
            axis.title.x=element_text(size=20),
            axis.title.y=element_text(size=20), 
            panel.spacing = unit(0.5, "lines"), 
            legend.position="none",
            legend.text=element_blank())+
    xlab("Average gene-gene correlation")+
    ylab("Average gene-cluster correlation"))

```

We can also construct a gene network based on the spatial vector for the
genes.
### Gene network
```{r gene_graph, warning=FALSE, message=FALSE, fig.width=6, fig.height=6}

vector_graph= igraph::graph_from_adjacency_matrix(cor_gene_mt,
                                                mode = "undirected", 
                                                weighted = TRUE, 
                                                diag = FALSE)

vector_graph=igraph::simplify(igraph::delete.edges(vector_graph, 
                        E(vector_graph)[abs(E(vector_graph)$weight) <= 0.7]))

layout=igraph::layout_with_kk(vector_graph)

# Plot the graph
ggraph::ggraph(vector_graph, layout = layout) +
    geom_edge_link(aes(edge_alpha = weight), show.legend = FALSE) +
    geom_node_point(color = "lightblue", size = 5) +
    geom_node_text(aes(label = name), 
                    vjust = 1, hjust = 1,size=2,color="orange", repel = TRUE) 

```

# Linear relationship between markers and clusters

We assume that the relationship between a marker gene vector its cluster
spatial vector is linear. \n

Here are several genes and their annotation from the panel. \n

|         |                            |
|---------|----------------------------|
| Gene    | Annotation                 |
| MLPH    | Breast glandular cells     |
| CXCR4   | T cells                    |
| MYLK    | Breast myoepithelial cells |
| CLEC14A | Adipocytes                 |
| LUM     | Fibroblasts                |

```{r linear vector to vector plot, fig.height=2, fig.width=8}

genes_lst = c("MLPH","CXCR4","MYLK","CLEC14A","LUM")

for (i_cluster in c("c1","c3","c5","c6","c8")){
    cluster_vector=rep1_sq10_vectors$cluster_mt[,i_cluster]
    
    data_vis=as.data.frame(cbind("cluster", cluster_vector,
                            rep1_sq10_vectors$gene_mt[, genes_lst]))
    
    colnames(data_vis)=c("cluster","cluster_vector",genes_lst)
    data_vis=reshape2::melt(data_vis,variable.name = "genes",
                            value.name = "gene_vector",
                            id= c("cluster","cluster_vector" ))
    data_vis$cluster_vector=as.numeric(data_vis$cluster_vector)
    data_vis$genes=factor(data_vis$genes)
    data_vis$gene_vector=as.numeric(data_vis$gene_vector)
    
    plot(ggplot(data = data_vis, 
            aes(x= cluster_vector, y=gene_vector))+
            geom_point()+
            facet_wrap(~genes,scales = "free_y", ncol=10)+
            theme_bw()+
            theme(legend.title=element_blank(),
                axis.text.y = element_text(size=6),
                axis.text.x = element_text(size=6,angle=0),
                axis.title.x=element_text(size=10),
                axis.title.y=element_text(size=10), 
                panel.spacing = unit(0.5, "lines"), 
                legend.position="none",
                legend.text=element_blank(),
                strip.text = element_text(size = rel(1)))+
                xlab(paste(i_cluster," - cluster vector", sep=""))+
                ylab("gene vector"))
}

```

# Scenario 1: one sample

A straightforward approach to identifying genes that exhibit a linear 
correlation with cluster vectors involves computing the Pearson correlation 
for each gene with every cluster. To assess the statistical significance 
of these correlations, the `compute_permp()` function can be used 
to perform permutation testing, generating a p-value for every pair of gene 
cluster and cluster vector. 

## Correlation-based method

### Permutation p-value

```{r rep1 permutation}

w_x =  c(min(floor(min(rep1_sub$trans_info$x)),
            floor(min(rep1_clusters$x))), 
        max(ceiling(max(rep1_sub$trans_info$x)),
            ceiling(max(rep1_clusters$x))))
w_y =  c(min(floor(min(rep1_sub$trans_info$y)),
            floor(min(rep1_clusters$y))), 
        max(ceiling(max(rep1_sub$trans_info$y)),
            ceiling(max(rep1_clusters$y))))
set.seed(seed_number)
perm_p = compute_permp(data=rep1_sub,
                        cluster_info=rep1_clusters, 
                        perm.size=1000,
                        bin_type="square",
                        bin_param=c(10,10),
                        all_genes=row.names(rep1_sub$cm),
                        correlation_method = "pearson", 
                        n.cores=1, 
                        correction_method="BH",
                        w_x=w_x ,
                        w_y=w_y)

perm_res = as.data.frame(perm_p$perm.pval.adj)
head(perm_res)
```

### Visualise top3 marker genes detected by permutation
Genes with a significant adjusted p-value are considered as marker genes for the 
correpsonding cluster. We can rank the marker genes by the observed correlation
and plot the transcript detection coordinates for the top three marker genes
for every cluster. 
```{r rep1 permutation vis, fig.width=6, fig.height=2}
res_df_1000=as.data.frame(perm_p$perm.pval.adj)
res_df_1000$gene=row.names(res_df_1000)
cluster_names = unique(as.character(rep1_clusters$cluster))
for (cl in cluster_names){
    perm_sig = res_df_1000[res_df_1000[,cl]<0.05,]
    perm_sig = as.data.frame(cbind("adj_pval" = perm_sig[,cl],
                     "obs_corr" =perm_p$obs.stat[row.names(perm_sig),cl] ))
    perm_sig = perm_sig[order(perm_sig[,"obs_corr"],decreasing = TRUE),]
    inters = row.names(perm_sig)[1:min(3, nrow(perm_sig))]
    inters_df = as.data.frame(cbind(gene=inters,
                            text=paste(inters,
                            round(as.numeric(perm_sig[inters,"obs_corr"]),3), 
                            sep="\n")))
    data_vis =rep1_sub$trans_info[rep1_sub$trans_info$feature_name %in% inters,
                            c("x","y","feature_name")]
    data_vis$text = inters_df[match(data_vis$feature_name,inters_df$gene),
                                "text"]
    data_vis$text = factor(data_vis$text, levels=inters_df$text)
    p1<-ggplot(data = data_vis,
                aes(x = x, y = y))+
                geom_point(size=0.01,color="black")+
                facet_wrap(~text,ncol=10, scales="free")+
                scale_y_reverse()+
                guides(fill = guide_colorbar(height= unit(5, "cm")))+
                theme(strip.text = element_text(size = rel(1)),
                    axis.line=element_blank(),
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="right",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())
    cl_pt<-ggplot(data = rep1_clusters[rep1_clusters$cluster==cl, ],
                    aes(x = x, y = y, color=cluster))+
                    geom_point(position=position_jitterdodge(jitter.width=0, 
                                            jitter.height=0), size=0.2)+
                    facet_wrap(~cluster)+
                    scale_y_reverse()+
                    theme_classic()+
                    scale_color_manual(values = "black")+
                    theme(axis.line=element_blank(),
                    strip.text = element_text(size = rel(1)), 
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="none",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())+
                    xlab("")+
                    ylab("")
        lyt <- cl_pt | p1
        layout_design <- lyt + patchwork::plot_layout(widths = c(1,4)) 
        print(layout_design)
}
```

The other method to identify linearly correlated genes for each cluster is
to construct a linear model for each gene. We can use the `lasso_markers`
function to get the most relevant cluster label for every gene.
## Linear model method

### Include negative control genes as background noise

We will create spatial vectors for negative control genes and include
them as background noise "clusters".
```{r rep1 lasso_markers with background}

rep1_nc_vectors = create_genesets(data_lst=list("rep1"= rep1_neg),
                                    name_lst=list(probe=rep1_neg$probe, 
                                                codeword=rep1_neg$codeword), 
                                    bin_type="square",
                                    bin_param=c(10, 10), 
                                    w_x=w_x, w_y=w_y)

set.seed(seed_number)
rep1_lasso_with_nc = lasso_markers(gene_mt=rep1_sq10_vectors$gene_mt,
                                    cluster_mt = rep1_sq10_vectors$cluster_mt,
                                    sample_names=c("rep1"),
                                    keep_positive=TRUE, 
                                    coef_cutoff=0.2,
                                    background=rep1_nc_vectors,n_fold = 5)

rep1_top_df_nc = rep1_lasso_with_nc$lasso_top_result
head(rep1_top_df_nc)
```


### Visualise the marker genes for each cluster
We can rank the marker genes by its linear model coefficient to the cluster
ans plot the transcript detection coordinates for the top three marker genes
for every cluster. 
```{r lasso-rep1 vis, fig.width=6, fig.height=2}
cluster_names = unique(as.character(rep1_top_df_nc$top_cluster))
for (cl in setdiff(cluster_names,"NoSig")){
    inters=rep1_top_df_nc[rep1_top_df_nc$top_cluster==cl,"gene"]
    rounded_val=signif(as.numeric(rep1_top_df_nc[inters,"glm_coef"]),
                          digits = 3)
    inters_df = as.data.frame(cbind(gene=inters, value=rounded_val))
    inters_df$value = as.numeric(inters_df$value)
    inters_df=inters_df[order(inters_df$value, decreasing = TRUE),]
    inters_df$text= paste(inters_df$gene,inters_df$value,sep=": ")
    
    if (length(inters > 0)){
        inters_df = inters_df[1:min(3, nrow(inters_df)),]
        inters = inters_df$gene
        iters_rep1= rep1_sub$trans_info$feature_name %in% inters
        vis_r1 =rep1_sub$trans_info[iters_rep1,
                                c("x","y","feature_name")]
        vis_r1$value = inters_df[match(vis_r1$feature_name,inters_df$gene),
                                "value"]
        #vis_r1=vis_r1[order(vis_r1$value,decreasing = TRUE),]
        vis_r1$text_label= paste(vis_r1$feature_name,
                                    vis_r1$value,sep=": ")
        vis_r1$text_label=factor(vis_r1$text_label, levels = inters_df$text)
        vis_r1$sample="rep1"
        p1<- ggplot(data = vis_r1,
                    aes(x = x, y = y))+ 
                    geom_point(size=0.01,color="black")+
                    facet_wrap(~text_label,ncol=10, scales="free")+
                    scale_y_reverse()+
                    guides(fill = guide_colorbar(height= unit(5, "cm")))+
                    theme(strip.text = element_text(size = rel(0.7)), 
                    axis.line=element_blank(),
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="right",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())
       
        
        cl_pt<-ggplot(data = rep1_clusters[rep1_clusters$cluster==cl, ],
                    aes(x = x, y = y, color=cluster))+
                    geom_point(position=position_jitterdodge(jitter.width=0, 
                                            jitter.height=0), size=0.2)+
                    facet_wrap(~cluster)+
                    scale_y_reverse()+
                    theme_classic()+
                    scale_color_manual(values = "black")+
                    theme(axis.line=element_blank(),
                    strip.text = element_text(size = rel(1)), 
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="none",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())+
                    xlab("")+
                    ylab("")
        lyt <- cl_pt | p1
        layout_design <- lyt + patchwork::plot_layout(widths = c(1,4)) 
        print(layout_design)
}}
```


# Scenario 2: multiple samples
Load the replicate 2 from sample 1. 
```{r load rep2 data}
data(rep2_sub, rep2_clusters, rep2_neg)
rep2_clusters$cluster=factor(rep2_clusters$cluster,
                            levels=paste("c",1:8, sep=""))
rep1_clusters$cells = paste(row.names(rep1_clusters),"_1",sep="")
rep2_clusters$cells =paste(row.names(rep2_clusters),"_2",sep="")
rep_clusters = rbind(rep1_clusters,rep2_clusters)
rep_clusters$cluster=factor(rep_clusters$cluster,
                            levels=paste("c",1:8, sep=""))
table(rep_clusters$cluster)
```

### Visualise the clusters
We can plot the coordiantes of cells for every cluster in every replicate 
```{r tworep cluster vis, fig.height=2,  fig.width=8, warning=FALSE}

ggplot(data = rep_clusters,
        aes(x = x, y = y, color=cluster))+
        geom_point(position=position_jitterdodge(jitter.width=0, 
                                                jitter.height=0),size=0.1)+
        facet_grid(sample~cluster)+
        scale_y_reverse()+
        theme_classic()+
        scale_color_manual(values = c("#FC8D62","#66C2A5" ,"#8DA0CB","#E78AC3",
                                "#A6D854","skyblue","purple3","#E5C498"))+
        guides(color=guide_legend(title="cluster", nrow = 1,
        override.aes=list(alpha=1, size=7)))+
        theme(
            axis.line=element_blank(),
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.title.x=element_blank(),
            axis.title.y=element_blank(),
            panel.background=element_blank(),
            panel.border=element_blank(),
            panel.grid.major=element_blank(),
            panel.grid.minor=element_blank(),
            plot.background=element_blank(),
            legend.text = element_text(size=10),
            legend.position="none",
            legend.title = element_text(size=10),
            strip.text = element_text(size = rel(1)))+
        xlab("")+
        ylab("")
```

When we have multiple replicates in the dataset, we can find marker
genes by providing additional sample information as the input for the function
`lasso_markers`.

```{r tworep no background}
w_x =  c(min(floor(min(rep1_sub$trans_info$x)),
            floor(min(rep2_sub$trans_info$x)),
            floor(min(rep_clusters$x))), 
        max(ceiling(max(rep1_sub$trans_info$x)),
            ceiling(max(rep2_sub$trans_info$x)),
            ceiling(max(rep_clusters$x))))
w_y =  c(min(floor(min(rep1_sub$trans_info$y)),
            floor(min(rep2_sub$trans_info$y)),
            floor(min(rep_clusters$y))), 
        max(ceiling(max(rep1_sub$trans_info$y)),
            ceiling(max(rep2_sub$trans_info$y)),
            ceiling(max(rep_clusters$y))))

all_genes = row.names(rep1_sub$cm)

# get spatial vectors
two_rep_vectors = get_vectors(data_lst= list("rep1"=rep1_sub, 
                                            "rep2" = rep2_sub),
                            cluster_info = rep_clusters, bin_type="square",
                            bin_param=c(10, 10), all_genes = all_genes , 
                            w_x=w_x, w_y=w_y)

two_rep_nc_vectors = create_genesets(data_lst=list("rep1" = rep1_neg, 
                                                    "rep2" = rep2_neg),
                                        name_lst=list(probe=rep1_neg$probe, 
                                                    codeword=rep1_neg$codeword),
                                        bin_type="square",
                                        bin_param=c(10,10), 
                                        w_x=w_x, w_y=w_y)
set.seed(seed_number)
two_rep_lasso_with_nc = lasso_markers(gene_mt=two_rep_vectors$gene_mt,
                                    cluster_mt = two_rep_vectors$cluster_mt,
                                    sample_names=c("rep1","rep2"),
                                    keep_positive=TRUE, 
                                    coef_cutoff=0.2,
                                    background=two_rep_nc_vectors,n_fold = 5)
tworep_res=two_rep_lasso_with_nc$lasso_top_result
tworep_res$celltype = rep_clusters[match(tworep_res$top_cluster,
                                            rep_clusters$cluster),"anno"]
head(tworep_res)
```

### Visualise the top3 marker genes for each cluster
```{r top3 marker gene, fig.height=3, fig.width=8}
all_celltypes = c("Tumor", "Myoepithelial", "DCIS", "T_Cells", "Stromal",
                    "Macrophages", "Endothelial", "B_Cells"  )
for (cl in all_celltypes){
    inters=tworep_res[tworep_res$celltype==cl,"gene"]
    rounded_val=signif(as.numeric(tworep_res[inters,"glm_coef"]),
                          digits = 3)
    inters_df = as.data.frame(cbind(gene=inters, value=rounded_val))
    inters_df$value = as.numeric(inters_df$value)
    inters_df=inters_df[order(inters_df$value, decreasing = TRUE),]
    inters_df$text= paste(inters_df$gene,inters_df$value,sep=": ")
    
    
    if (length(inters > 0)){
        inters_df = inters_df[1:min(3, nrow(inters_df)),]
        inters = inters_df$gene
        iters_rep1= rep1_sub$trans_info$feature_name %in% inters
        vis_r1 =rep1_sub$trans_info[iters_rep1,
                                c("x","y","feature_name")]
        vis_r1$value = inters_df[match(vis_r1$feature_name,inters_df$gene),
                                "value"]
        vis_r1=vis_r1[order(vis_r1$value,decreasing = TRUE),]
        vis_r1$text_label= paste(vis_r1$feature_name,
                                    vis_r1$value,sep=": ")
        vis_r1$text_label=factor(vis_r1$text_label)
        vis_r1$sample="rep1"
        iters_rep2= rep2_sub$trans_info$feature_name %in% inters
        vis_r2 =rep2_sub$trans_info[iters_rep2,
                                c("x","y","feature_name")]
        vis_r2$value = inters_df[match(vis_r2$feature_name,inters_df$gene),
                                "value"]
        vis_r2=vis_r2[order(vis_r2$value, decreasing = TRUE),]
        vis_r2$text_label= paste(vis_r2$feature_name,
                                vis_r2$value,sep=": ")
        vis_r2$text_label=factor(vis_r2$text_label)
        vis_r2$sample="rep2"
        p1<- ggplot(data = vis_r1,
                    aes(x = x, y = y))+ 
                    geom_point(size=0.01,color="black")+
                    facet_wrap(sample~text_label,ncol=10, scales="free")+
                    scale_y_reverse()+
                    guides(fill = guide_colorbar(height= unit(5, "cm")))+
                    theme(strip.text = element_text(size = rel(0.7)), 
                    axis.line=element_blank(),
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="right",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())
        p2<- ggplot(data = vis_r2,
                    aes(x = x, y = y))+ 
                    geom_point(size=0.01,color="black")+
                    facet_wrap(sample~text_label,ncol=10, scales="free")+
                    scale_y_reverse()+
                    guides(fill = guide_colorbar(height= unit(5, "cm")))+
                    theme(strip.text = element_text(size = rel(0.7)), 
                    axis.line=element_blank(),
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="right",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())
        
        cl_pt<-ggplot(data = rep_clusters[rep_clusters$anno==cl, ],
                    aes(x = x, y = y, color=cluster))+
                    geom_point(position=position_jitterdodge(jitter.width=0, 
                                            jitter.height=0), size=0.2)+
                    facet_grid(sample~cluster)+
                    scale_y_reverse()+
                    theme_classic()+
                    scale_color_manual(values = "black")+
                    theme(axis.line=element_blank(),
                    strip.text = element_text(size = rel(1)), 
                    axis.text.x=element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks=element_blank(),
                    axis.title.x=element_blank(),
                    axis.title.y=element_blank(),
                    legend.position="none",
                    panel.background=element_blank(),
                    panel.border=element_blank(),
                    panel.grid.major=element_blank(),
                    panel.grid.minor=element_blank(),
                    plot.background=element_blank())+
                    xlab("")+
                    ylab("")
        lyt <- cl_pt | (p1 / p2) 
        layout_design <- lyt + patchwork::plot_layout(widths = c(1,4)) 
        print(layout_design)
}}
```

```{r session info}
sessionInfo()
```
